---
title: "Introduction to BLSloadR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BLSloadR-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(BLSloadR)
```

## Overview

**BLSloadR** is a packages designed to streamline access to the time series database downloads from the [U.S. Bureau of Labor Statistics](https://www.bls.gov/), made available at <https://download.bls.gov/pub/time.series/>. It is focused on accessing series that are frequently used by states to get state-level estimates, but includes the `load_bls_dataset()` function to provide generalized access to the other databases at this website.

## Basic Usage

The primary functions in this package all begin with get\_ and are listed below:

-   `get_ces()` - This accesses data from the Current Employment Statistics (CES) program at the state and metropolitan area levels. This provides employer-based estimates of employment, wages, and hours worked. This is the "SM" database.

-   `get_national_ces()` - This accesses national data from the CES program at the national level, which does not include state-level breakouts. This is the "CE" database.

-   `get_laus()` - This accesses data from the Local Area Unemployment Statistics (LAUS) program at a regional, state, and several sub-state levels. This is a localized version of the Current Population Survey (CPS) which is used to drive household-based estiamtes of employment and unemployment. This is the "LA" database. Note that because of the volume of data here, there are several different geographies that may be specified to pull the appropriate data file from BLS.

-   `get_oews()` - This access the Occupational Employment and Wage Statistics (OEWS) data. This data provides survey-based estimates of employmen and wages by occupation at state and sub-state levels. This is the "OE" database. Note that only current-year data is available for OEWS in this database, as it is not built as a time series.

-   `get_salt()` - This data is not actually loated within the time.series folder, but instead is sourced from <https://www.bls.gov/lau/stalt.htm>. These *Alternative Measures of Labor Underutilization for States* are 12-month averages built from CPS data which provide more expansive or restrictive definitions of unemployment to measure the labor force, known as U1 through U6. This function also includes the optional geometry argument. If set to TRUE, this will use `tigris::states()` and `tigris::shift_geometry()` to provide state polygons for convenient mapping of the output.

These optional helper functions can aid the user of this package by providing ways to summarize and explore all the time.series databases.

-   `bls_overview()` - this function utilizes the standard structure of the time.series databases, which has a simple text file explaining the database structure that always follows the structure *id.txt* where *id* is the two-character database identification code.

-   `load_bls_dataset()` - this function attempts to read and join all the relevant files in a BLS database, and will sometimes prompt the user for additional input. For example, many databases have multiple data files available (such as "AllItems" and "Current") and may have old series files as well (to manage historical coding changes). Because these joins are performed automatically, the object returned by this function is a more robust diagnostic object included the joined data table as well as information about the joins. **Use Caution!** BLS data structures are not always consistent. There may be anomalies in the structure of individual databases, such as missing column headers, that will degrade the ability of this function to read the data.

## Additional Diagnostics

Within each of the core functions is an argument that allows you to receive more robust diagnostic information about the loading process. setting `return_diagnostics = TRUE` will return a bls_data_collection object, which can be used to get more information about the loading process. For `load_bls_dataset()` you should set `return_full = TRUE`.
